Tesseroids: command-line tools for gravity field forward modeling in spherical coordinates

### 2014-04-01 manuscript.tex
Section on comparison with shell
Finished most of it and updated the figure to include the height of the
computation point.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nam eu dolor pretium,
egestas mauris sed, dapibus quam. Duis hendrerit mollis nunc a consequat. Nulla
et sem consectetur, interdum velit eget, aliquam ipsum. Praesent sagittis
tortor diam, sed ultrices magna ullamcorper vitae. Proin vitae orci augue.
Morbi dictum ligula gravida sem malesuada facilisis. Mauris nibh metus, cursus
eget imperdiet vitae, pretium at lorem. Praesent nisi mauris, pretium ut risus
fermentum, egestas tincidunt nibh. Mauris nulla orci, consequat eu pharetra
non, mattis ut urna. Mauris facilisis orci eros. Nam mattis non magna iaculis
consectetur. Morbi sodales dolor vitae felis sagittis, eget faucibus turpis
convallis. Nullam malesuada, mauris et ultricies rutrum, odio nulla gravida
nunc, ac volutpat eros lectus eget lacus. Integer venenatis velit vel justo
pellentesque, quis molestie sem vestibulum.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology}

@@

The gravitational potential
of a homogeneous spherical half-shell
at a computation point P
located at a height $h$
from the center of the shells top face
(Figure~\ref{fig:shell})
is

\begin{equation}
    V(r) = 2\pi G \rho \left[ \dfrac{l^3 + {r'}^3}{3r} - 0.5 {r'}^2 \right]
           \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:halfshell-pot}
\end{equation}

\noindent
where $\rho$ is the density,
$l = \sqrt{r^2 + {r'}^2}$,
$r$ is the radial coordinate of the observation point,
and
$r_1$ and $r_2$ are the bottom and top of the spherical shell,
respectively.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/spherical-shell}
    \caption{This is a figure caption.}
    \label{fig:shell}
\end{figure}

This is a reference to Figure~\ref{fig:shell}.

The $g_z$ component of the gravitational attraction for this shell is the
radial derivative of the potential.

The sign should be inverted because our convention is that z points down
(toward the center of the Earth)

\begin{equation}
    g_z(r) = -\dfrac{\partial V}{\partial r} = 2\pi G \rho \left[ \dfrac{l^3 +
            {r'}^3}{3r^2} - l \right] \Biggr \rvert_{r'=r_1}^{r'=r_2}
    \label{eq:halfshell-gz}
\end{equation}

The gravity gradient tensor are the second derivatives of $V$:

\begin{equation}
    g_{xx}(r) = g_{yy}(r) = -\dfrac{1}{2} g_{zz}
\end{equation}

\begin{equation}
    g_{xy}(r) = g_{xz}(r) = g_{yz}(r) = 0
\end{equation}

\begin{equation}
    g_{zz}(r) = 2\pi G \rho \left[ 2\dfrac{l^3 + {r'}^3}{3r^3} - \dfrac{l}{r} +
    \dfrac{r}{l} \right] \Biggr \rvert_{r'=r_1}^{r'=r_2}
\end{equation}

@@

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nam eu dolor pretium,

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

%\begin{figure}
    %\centering
    %\subfigure[]{
        %\includegraphics[width=0.4\textwidth]{figs/sub1.png}\label{fig:sub1}}
    %\subfigure[]{
        %\includegraphics[width=0.4\textwidth]{figs/sub2.png}\label{fig:sub2}}
  %\caption{This is a figure caption.}
  %\label{fig:composite}
%\end{figure}

%This Figure \ref{fig:composite} is made up of Figure \ref{fig:sub1} and
%\ref{fig:sub2}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgments}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{seg}
\bibliography{references}

>>>>

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nam eu dolor pretium,
egestas mauris sed, dapibus quam. Duis hendrerit mollis nunc a consequat. Nulla
et sem consectetur, interdum velit eget, aliquam ipsum. Praesent sagittis
tortor diam, sed ultrices magna ullamcorper vitae. Proin vitae orci augue.
Morbi dictum ligula gravida sem malesuada facilisis. Mauris nibh metus, cursus
eget imperdiet vitae, pretium at lorem. Praesent nisi mauris, pretium ut risus
fermentum, egestas tincidunt nibh. Mauris nulla orci, consequat eu pharetra
non, mattis ut urna. Mauris facilisis orci eros. Nam mattis non magna iaculis
consectetur. Morbi sodales dolor vitae felis sagittis, eget faucibus turpis
convallis. Nullam malesuada, mauris et ultricies rutrum, odio nulla gravida
nunc, ac volutpat eros lectus eget lacus. Integer venenatis velit vel justo
pellentesque, quis molestie sem vestibulum.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology}

@@

The gravitational potential,
gravitational attraction,
and
Marussi (gravity gradient) tensor
caused by the homogeneous spherical half-shell
shown in Figure~\ref{fig:shell}
at a computation point
$P = (\lambda=90^\circ, \phi=0^\circ, r=r_2+h)$
are

\begin{equation}
    V(h) = 2\pi G \rho \left[
        \dfrac{l^3 + {r'}^3}{3(r_2 + h)} - 0.5 {r'}^2 \right]
         \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellpot}
\end{equation}
\begin{equation}
    g_z(h) = 2\pi G \rho \left[
        \dfrac{l^3 + {r'}^3}{3(r_2 + h)^2} - l \right]
        \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellgz}
\end{equation}
\begin{equation}
    g_{xx}(h) = g_{yy}(h) = -\dfrac{1}{2} g_{zz}(h),
\end{equation}
\begin{equation}
    g_{xy}(h) = g_{xz}(h) = g_{yz}(h) = 0,
\end{equation}

\begin{equation}
    g_{zz}(h) = 2\pi G \rho \left[
        2\dfrac{l^3 + {r'}^3}{3(r_2 + h)^3}
        - \dfrac{l}{r_2 + h} + \dfrac{r_2 + h}{l}
        \right]
        \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellgzz}
\end{equation}

\noindent
where $\rho$ is the density,
$l = \sqrt{(r_2 + h)^2 + {r'}^2}$,
and
$r_1$ and $r_2$ are radial coordinates of
the bottom and top of the spherical shell,
respectively.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/spherical-shell}
    \caption{A spherical half-shell
        spanning the ranges $0^{\circ} \le \lambda \le 180^\circ$,
        $-90^\circ \le \phi \le 90^\circ$,
        and $r_1 \le r \le r_2$,
        the corresponding tesseroid model,
        the computation point $P$
        at $(\lambda=90^\circ,\phi=0^\circ,r=r_2+h)$,
        and its local coordinate system $(x, y, z)$.
    }
    \label{fig:shell}
\end{figure}

For our comparison,
we used a spherical half-shell with
$\rho=2.7\ g.cm^{-3}$,
$r_1=6378.137\ km$ (mean Earth radius),
and
$r_2 = r_1 + 50\ km$.
We discretized the half-shell
into tesseroids of equal size
and calculated the gravitational effects of
the half-shell
(using equations \ref{eq:shellpot}-\ref{eq:shellgzz})
and the corresponding tesseroid model
at various heights.
These calculations were repeated
for varying values of
the distance/size ratio $R$.
This analysis
allows us to establish
the optimal value of $R$
that achieves a desired error level.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgments}

@@

\bibliographystyle{seg}
\bibliography{references}
